One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-10-12 doc/genesetdp.tex
Edits

% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics. For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
The methods that are used for network randomization, while computationaly efficient, are often mentally untractable as they include a big number of steps that are subject to complex constrains.
As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow contrains that depends on previous link assignments.
While this results in a network that is both randomized and contain some preserved desired topological features, these nested constrains make it very hard to verify that the resulting network is representative. In theory there is a space of all possible networks that could be made up from a certain node degree distribution, however, it is not clear that the  second-order conservation method samples evenly from that distribution.  And if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose the statistical test.

Users of NEA typically want to test if there is a functional association between a query set of genes and a pathway\cite{alexeyenko2012network}. However, we argue that the network pertubation theory used in most flavors of NEA is a red herring. The typical user have used an experimental or computational method to derive a set of query genes that the user want to know if it is random, or if it is enriched in repect to a pathway.  In this question, the network is not relevant. A network may help to increase the sensitivity of ORA methods, but it is better left out of the null formation of the null hypothesis. We argue that such test is naturally implemented as tests of the query set, i.e. is the query set a randomly selected set of genes or is there an over representation of well-connected genes in the query set?

@@

>>>>

% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics. For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
The methods that are used for network randomization, are often mentally untractable as they include a big number of steps that are subject to complex constrains.
As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow contrains that depends on previous link assignments.
While this results in a network that is both randomized and contain some preserved desired topological features, these nested constrains make it very hard to verify that the resulting network is representative. In theory there is a space of all possible networks that could be made up from a certain node degree distribution, however, it is not clear that the  second-order conservation method samples evenly from that distribution.  And, if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose the statistical test.

We argue that the network pertubation theory used in most flavors of NEA is a red herring. Users of NEA do not differ from users of ORA they want to test an association between a query set of genes and a pathways.  The typical user have used an experimental or computational method to derive a set of query genes that the user want to know if it is random, or if it is enriched in repect to a pathway.  In this question, the introduction of a  network is not relevant. A network may help to increase the sensitivity of ORA methods, but it is better left out of the null hypothesis. We argue that such test is naturally implemented as tests of the query set, i.e. is the query set a randomly selected set of genes or is there an over representation of well-connected genes in the query set?

@@

In effect, NEA is an attempt to increase the sensitivity of ORA. The standard user of NEA would typically have identified a set of query genes which she wants to evaluate for over representation in respect to a pathway database. This goal is the same as for users of ORA. Hence the null model should be of the same nature for NEA and ORA. GeneSetDP and GeneSetMC are two algorithms that proves that one in NEA can test the uniqness of the querry set in respect to a pathway, just as done in ORA.
